---
title: "Class19 Mini Project: Investigating Pertussis Resurgence"
author: "Vincent Chau"
format: gfm
---
## Investigating pertussis cases by year

```{r}
cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L),
         Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617)
       )
```


Make a plot of the number of Pertussis cases per year from this CDC data to investigate

>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(datapasta)
library(ggplot2)
baseplot <- ggplot(cdc) +
  aes(x=Year, y=Cases) +
  geom_line() +
  labs(title = "Number of reported Petussis Cases per ")
baseplot
```
>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
baseplot + geom_vline(xintercept = 1946, linetype=2, color="blue") + geom_vline(xintercept = 1996, linetype=2, color="red")

```

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

It is clear from the CDC data that pertussis cases are once again increasing. For example, we can see that in 2012 the CDC reported 48,277 cases of pertussis in the United States. This is the largest number of cases reported since 1955, when 62,786 cases were reported. The pertussis field has several hypotheses for the resurgence of pertussis including (in no particular order): 1) more sensitive PCR-based testing, 2) vaccination hesitancy 3) bacterial evolution (escape from vaccine immunity), 4) waning of immunity in adolescents originally primed as infants with the newer aP vaccine as compared to the older wP vaccine.

## Getting Data from CMI-PB

The CMI-PB resource is studying and making available data on the immune response to Petussis vaccination.

It mostly returns JSON format data that we need to process and conert into something useable in R.

```{r}
# Allows us to read, write and process JSON data
library(jsonlite)
```
```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)
```

```{r}
head(subject, 3)
```

>Q4. How may aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```
>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```
>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
## Side-Note: Working with dates

To help ease the pain of working with dates we can use `library(lubridate)`.

```{r}
library(lubridate)
```

```{r}
today()
```
How many days have passed since new year 2000

```{r}
today() - ymd("2000-01-01")
```
What is this in years?

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```
>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <- today() - ymd(subject$year_of_birth)
subject$age
```
Now we can select or filter for "aP" and find the average age - convert into years.

```{r}
library(dplyr)
ap <- subject %>% filter(infancy_vac == "aP")
round( summary( time_length( ap$age, "years")))
wp <- subject %>% filter(infancy_vac == "wP") 
summary(time_length(wp$age, "years"))
```
>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```
>Q9. With the help of a faceted boxplot (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2)
```

## Joining multiple tables
## Ab titer data

```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api//specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", simplifyVector = TRUE) 
```


Quick peak at the start of this specimen data

```{r}
head(specimen)
```

>Q9. Complete the code to join ``specimen` and `subject` tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

```{r}
dim(subject)
```

>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

>Q12. What do you notice about the number of visit 8 specimens compared to other visits?

```{r}
table(abdata$visit)
```

## Examine IgG1 Ab titer levels

Exclude the small number of visit 8 entries.

```{r}
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```

>Q13. Complete the following code to make a summary boxplot of Ab titer levels for all antigens:

Make a summary plot of antigen levels.
```{r}
ggplot(ig1) + 
  aes(MFI, antigen) +
  geom_boxplot()
# facet by aP or wP
```

```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() +
  facet_wrap(vars(infancy_vac))
```

How do these change over time - one viz that may be helpful is to facet the adobe plot by visit number.

```{r}
ggplot(ig1) + 
  aes(MFI, antigen) +
  geom_boxplot()+
  facet_wrap(vars(visit), nrow=2)
```

